---
title: "PA1_template"
author: "Raghav Luthra"
date: "28 August 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

Loading the required packages:  

``` {r load, message=FALSE , warning=FALSE}

  library(ggplot2)
  library(lubridate) 
  library(dplyr)
  library(gridExtra)
  
```

## Reading the file and understanding it

```{r read}
  activity <- read.csv("activity.csv")

  str(activity)
  
  sum(is.na(activity$steps))
  
```

As described in introduction, the dataset consists of 3 columns viz. :  

- Date : The date of the experiment
- Interval : The 5 minute interval's starting point
- Steps : Number of steps taken in that interval

Also, we see that there are `r sum(is.na(activity$steps))` NA values and the date column is also a factor. So, we need to do something about the NA values and convert factors into proper date formats. However, this will be done later as we progress through all the questions

## What is mean total number of steps taken per day?

```{r hist_na,fig.width=8}

  ## Subsetting the non-NA values
  activity_new <- activity[!is.na(activity$steps),]

  ## Converting date into characters to group them
  activity_new$date <- as.character(activity_new$date)

  ## Finding mean, median and total steps of each day
  summary_new <- summarise(group_by(activity_new,date), Mean_steps =mean(steps),  Median_steps=median(steps), Sum=sum(steps))

  summary_frame <- as.data.frame(summary_new)

  print(summary_frame)
  
  ## Converting them back to date format for plotting purpose
  summary_frame$date <- as.POSIXct(summary_frame$date)
   
  ggplot(data=summary_frame, aes(date,Sum)) + geom_bar(stat = "identity") + xlab("Date") + ylab("Total steps") + ggtitle("Total steps each day from Oct, 2012 to Nov, 2012")

```

## What is the average daily activity pattern?

```{r daily}

 ## Summarising by each 5 minute interval
 summary_time <- summarise(group_by(activity_new,interval),Mean_step=mean(steps))

 summary_time <- as.data.frame(summary_time)

 plot(summary_time$interval,summary_time$Mean_step,type = "l",col="brown",lwd=1.5, xlab = "Interval",ylab =   "Mean Step", main = "Time series of mean number of steps")
  
```

### Finding the interval with the maximum number of average steps

```{r max}

summary_time[summary_time$Mean_step==max(summary_time$Mean_step),]

```

The maximum number of average steps is 206.17 from time 08:35 AM to 08:40 AM.

## Inputting Missing Values

Here, the NA value for each interval 0,5,10 etc will be replaced by the mean value of steps taken by the human over the entire period of test in that particular interval.

``` {r missing-val}

interval_times <- unique(activity$interval)

n <- length(interval_times)

  for( i in 1 : n)
  {
    interval_data <- activity[activity$interval==interval_times[i],]
    
    mean_steps <- mean(interval_data$steps, na.rm=T)
    
    activity [activity$interval==interval_times[i] & is.na(activity$steps), "steps" ] <- mean_steps
  }

activity$steps <- round(activity$steps,3)

```


## Finding sum, mean, median number of steps for each day with new values

``` {r s-m-m,fig.width=8}

  activity$date <- as.character(activity$date)

  ## Finding mean, median and total steps of each day with all values
  summary <- summarise(group_by(activity,date), Mean_steps =mean(steps), Median_steps=median(steps),   Sum=sum(steps))

  summary <- as.data.frame(summary)
 
  print(summary)
  
  summary$date <- as.POSIXct(summary$date)
   
  ggplot(data=summary, aes(date,Sum)) + geom_bar(stat = "identity") + xlab("Date") + ylab("Total steps") + ggtitle("Total steps each day from Oct, 2012 to Nov, 2012")

```

## Clssifying the dates into weekday or weekend

``` {r weekday-weekend}

  activity$date <- as.POSIXlt(activity$date)

  ## 1: Sunday and 7: Saturday
  activity$type <- ifelse( wday(activity$date) == 1 | wday(activity$date) == 7 ,"weekend" ,"weekday")

  activity$type <- as.factor(activity$type)

```

## Are there differences in activity patterns between weekdays and weekends?

``` {r time-w, fig.width=8, fig.height=9}

  activity$date <- as.character(activity$date)

  ## Splitting data into weekend and weekday data
  split_data <- split(activity,activity$type)

  weekday_data <- split_data[[1]]

  weekend_data <- split_data[[2]]

  ## Finding mean steps for each interval on weekdays
  summary_day <- summarise(group_by(weekday_data,interval), Mean_steps =mean(steps))

  summary_day  <- as.data.frame(summary_day)

  ## Finding mean steps for each interval on weekends
  summary_end <- summarise(group_by(weekend_data,interval), Mean_steps =mean(steps))

  summary_end  <- as.data.frame(summary_end)

  g1 <- ggplot(data=summary_day,aes(interval,Mean_steps)) + geom_line(colour="blue") + xlab("Interval") + ylab("Mean Step") + ggtitle("Weekday")

  g2 <- ggplot(data=summary_end,aes(interval,Mean_steps)) + geom_line(colour="blue") + xlab("Interval") + ylab("Mean Step") + ggtitle("Weekend")

  ## Plotting them on the same canvas
  grid.arrange(g1,g2,ncol=1)

```
